---
title: "README"
author: "Jens Riis Baalkilde"
date: "December 11, 2024"
output: 
  md_document:
      variant: gfm
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


bmd
========================================================

<!-- badges: start -->
[![R-CMD-check](https://github.com/hreinwald/bmd/actions/workflows/r.yml/badge.svg)](https://github.com/hreinwald/bmd/actions/workflows/r.yml)
<!-- badges: end -->

This is an R package for benchmark dose (BMD) estimation, which expands upon the functionality of the drc package. 

This package is currently maintained by Jens Riis Baalkilde and Signe Marie Jensen, Department of Plant and Environmental Sciences, University of Copenhagen.

Authors: Signe M. Jensen, Christian Ritz and Jens Riis Baalkilde.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(drcData)
library(drc)
library(bmd)
```


Installation
-------------------------

Install the bmd package from GitHub. The bmd package works best with the drc package installed from GitHub as well.

```{r installation, eval=FALSE}
install.packages("devtools")
devtools::install_github("DoseResponse/drcData")
devtools::install_github("DoseResponse/drc")
devtools::install_github("hreinwald/bmd")
```

**Requirements:** R (>= 3.5). The package imports `drc`, `ggplot2`, `dplyr` and `stats`.

Key features
-------------------------

The bmd package is a natural extension of the drc package. Key features of the bmd package includes:

- Benchmark dose (BMD) estimation for quantal, continuous, count and ordinal data for a wide range of definitions of BMD.
  - A range of confidence intervals for BMD estimates, including Wald, Profile and Bootstrap intervals.
  - BMD estimation based on model averaging.
  - BMD estimation for multiple dose-response curves.
- Specialized model fitting for heterogeneous variance, ordinal and meta-analytic dose-response data.
- Isotonic regression-based BMD estimation.
- Statistical tests for trend and monotonicity (Williams, Shirley, Jonckheere, Bartholomew).
- Utility functions for plotting dose-response curves and BMD estimates.

Exported functions
-------------------------

### Core BMD estimation

| Function | Description |
|---|---|
| `bmd()` | BMD estimation for dose-response models fitted by `drc::drm()` |
| `bmdBoot()` | Bootstrap-based BMD with confidence intervals |
| `bmdMA()` | Model-averaged BMD estimation |
| `bmdMACurve()` | Model-averaged BMD curve helper |

### Specialized BMD estimation

| Function | Description |
|---|---|
| `bmdOrdinal()` | BMD for ordinal dose-response data |
| `bmdOrdinalMA()` | Model-averaged BMD for ordinal data |
| `bmdHetVar()` | BMD for models with heterogeneous variance |
| `bmdHetVarMA()` | Model-averaged BMD with heterogeneous variance |
| `bmdIso()` | BMD based on isotonic regression |
| `bmdIsoBoot()` | Bootstrap BMD based on isotonic regression |

### Dose-response model fitting

| Function | Description |
|---|---|
| `drmHetVar()` | Fit dose-response models with heterogeneous variance |
| `drmOrdinal()` | Fit ordinal dose-response models |
| `drmMMRE()` | Fit meta-analytic random-effects dose-response models |

### Plotting

| Function | Description |
|---|---|
| `qplotDrc()` | Plot dose-response curves using ggplot2 |
| `qplotBmd()` | Plot BMD estimates with confidence intervals |

### Statistical tests

| Function | Description |
|---|---|
| `trendTest()` | Trend tests (Williams, Shirley, Tukey) |
| `monotonicityTest()` | Monotonicity tests (Jonckheere, Bartholomew, Williams) |

### Utilities

| Function | Description |
|---|---|
| `BCa()` | Bias-corrected and accelerated bootstrap intervals |
| `PAV()` | Pool-Adjacent-Violators algorithm (isotonic regression) |
| `MACurve()` | Model-averaged curve computation |
| `getStackingWeights()` | Compute stacking weights for model averaging |

Example usage
-------------------------

In the following it is demonstrated how to use the functions in the bmd package on two data sets with a continuous response variable and one data set with a quantal response.

### Example 1 (secalonic)

First, the data set from the drcData package is loaded, and a dose-response model is fitted. 

```{r secalonic}
data("secalonic")
secalonic.LL.4 <- drm(rootl ~ dose, data = secalonic, fct = LL.4())
```

Users familiar with the drc package already know how straight-forward it is to plot dose-response models using the plot() function. The qplotDrc() function in the bmd package mimics the same functionality based on ggplot2. 

```{r secalonic_plot}
plot(secalonic.LL.4, main = "Secalonic model plotted by basic R plotting")

qplotDrc(secalonic.LL.4) + ggplot2::labs(title = "Secalonic model plotted by ggplot2")
```

Similar to the old method of plotting dose-response curves, qplotDrc() features several options for customisation.
```{r secalonic_plot_features}
qplotDrc(secalonic.LL.4, type = "all")
qplotDrc(secalonic.LL.4, type = "obs")
qplotDrc(secalonic.LL.4, type = "bars")
qplotDrc(secalonic.LL.4, type = "confidence")
```


#### Example 1: BMD estimation

A wide range of BMD definitions are implemented in the bmd package. In the following, the BMD and BMDL based on the "relative risk" definition recommended by EFSA, with a BMR = 10% are computed.

```{r secalonic_bmd}
bmd(secalonic.LL.4, bmr = 0.1, backgType = "modelBased", def = "relative")
```

The default interval is a Wald type confidence interval. If a profile likelihood confidence interval is preferred, this can be specified by the "interval" argument.

```{r secalonic_bmd_profile}
bmd(secalonic.LL.4, bmr = 0.1, backgType = "modelBased", def = "relative", interval = "profile")
```


The current state-of-the-art definition of the BMD is the hybrid definition. In the following, the BMD and BMDL based on the "excess risk" definition based on the hybrid method with BMR = 10%, and adverse background level set to the estimated background level minus 1SD is calculated.

```{r secalonic_bmd_hybrid}
bmd(secalonic.LL.4, bmr = 0.1, backgType = "hybridSD", backg = 1, def = "hybridExc")
```

For technical reasons, profile intervals for BMD estimates based on the hybrid method, are not currently available. 

#### Example 1: BMD plotting

The qplotBmd() function ca be used to plot BMD along with the dose-response curve.

```{r secalonic_bmd_plot}
qplotBmd(bmd(secalonic.LL.4, bmr = 0.1, backgType = "modelBased", def = "relative", display = FALSE)) # display = FALSE hides output from bmd function
```


#### Example 1: BMD estimation using MA

An additional set of dose-response models are fitted.

```{r}
secalonic.LL.3 <- drm(rootl ~ dose, data = secalonic, fct = LL.3())
secalonic.LN.3 <- drm(rootl ~ dose, data = secalonic, fct = LN.3())
secalonic.LN.4 <- drm(rootl ~ dose, data = secalonic, fct = LN.4())
secalonic.W1.3 <- drm(rootl ~ dose, data = secalonic, fct = W1.3())
secalonic.W1.4 <- drm(rootl ~ dose, data = secalonic, fct = W1.4())
secalonic.W2.3 <- drm(rootl ~ dose, data = secalonic, fct = W2.3())
secalonic.W2.4 <- drm(rootl ~ dose, data = secalonic, fct = W2.4())

secalonic.modelList <- list(secalonic.LL.3, secalonic.LL.4, secalonic.LN.3, secalonic.LN.4,
                            secalonic.W1.3, secalonic.W1.4, secalonic.W2.3, secalonic.W2.4)
```

BMD can now be estimated by MA by using the function bmdMA. Type of model weights need to be specified as well as MA type. 

```{r secalonic_bmdMA, warning=FALSE}
bmdMA(secalonic.modelList, modelWeights = "AIC", type = "Buckland", bmr = 0.1, backgType = "modelBased", def = "relative")

set.seed(123)
bmdMA(secalonic.modelList, modelWeights = "AIC", type = "bootstrap", bmr = 0.1, backgType = "modelBased", def = "relative", R = 500, progressInfo = FALSE)
bmdMA(secalonic.modelList, modelWeights = "AIC", type = "Buckland", bmr = 0.1, backgType = "hybridSD", def = "hybridExc", backg = 1)
```




### Example 2 (S.alba.comp, two dose-response curves)

An experiment was conducted where some pots were treated with the herbicide Bentazone, and other pots were treated with Glyphosate. A model with separate curves for each herbicide is fitted as follows.

```{r S.alba.comp, warning=FALSE}
data("S.alba.comp")
S.alba.comp.LL.4 <- drm(drymatter ~ dose, curveid = herbicide, data = S.alba.comp, fct = LL.4())
```

To plot the function, qplotDrc can be used:
```{r S.alba.comp_plot}
qplotDrc(S.alba.comp.LL.4)
qplotDrc(S.alba.comp.LL.4, col = TRUE, type = "confidence") +
  qplotDrc(S.alba.comp.LL.4, col = TRUE, type = "obs", add = TRUE)$obsLayer
```

This also illustrates how qplotDrc() can be layered to produce custom plots, by using the argument "add = TRUE", and choosing either the layer of observations ("obsLayer"), the layer with the dose-response curve ("curveLayer"), or the layer with the confidence band around the curves ("confBandLayer").

#### Example 2: BMD estimation.

BMD and BMDL can be estimated for the individual curves:

```{r S.alba.comp_bmd}
bmd(S.alba.comp.LL.4, bmr = 0.1, backgType = "modelBased", def = "relative")
bmd(S.alba.comp.LL.4, bmr = 0.1, backgType = "hybridSD", backg = 1, def = "hybridExc")
```


### Example 3 (acute.inh, binomial response)

20 animals were exposed to 4 doses of an unknown substance, 5 animals per dose. The number of dead animals for each dose were recorded. In the following, the data is loaded, and a 2-parameter is fitted to the data. Subsequently, a plot of the resulting dose-response curve is created using qplotDrc().

```{r acute.inh, warning=FALSE}
data("acute.inh")
acute.inh.LL.2 <- drm(num.dead/total ~ dose, weights = total, data = acute.inh, fct = LL.2(), type = "binomial")

qplotDrc(acute.inh.LL.2)
```




#### Example 3: BMD Estimation

For binomial response data, the "excess" and "additional" BMD definitions are available. When the background response is $0$, they are identical:

```{r acute.inh_bmd}
bmd(acute.inh.LL.2, bmr = 0.1, backgType = "modelBased", def = "additional")
bmd(acute.inh.LL.2, bmr = 0.1, backgType = "modelBased", def = "excess")
```



```{r acute.inh_bmd_plot}
qplotBmd(bmd(acute.inh.LL.2, bmr = 0.1, backgType = "modelBased", def = "additional", display = FALSE)) +
  ggplot2::scale_x_continuous(limits = c(0,2100), trans = scales::transform_pseudo_log(2000))
```


Vignettes
-------------------------

The package includes detailed vignettes covering all major features. After installation, browse them with:

```{r vignettes, eval=FALSE}
browseVignettes("bmd")
```

Available vignettes:

- **bmd Package** — Overview of BMD concepts and core functionality
- **Basic BMD Functions** — Usage examples for `bmd()` and related functions
- **Model Averaging** — Guide to `bmdMA()` and model-averaged BMD estimation
- **Specialized Models** — Ordinal, heterogeneous variance and meta-analytic models
- **Visualization** — Plotting with `qplotDrc()`, `qplotBmd()` and `plot.bmd()`
- **Utilities** — Helper functions (`PAV()`, `BCa()`, `expandBinomial()`, `bootDataGen()`)
- **Statistical Tests** — Trend and monotonicity tests (`trendTest()`, `monotonicityTest()`)


Citation
-------------------------

If you use the `bmd` package in your work, please cite:

> Jensen, S. M., Kluxen, F. M., Streibig, J. C., Cedergreen, N., & Ritz, C. (2020). bmd: an R package for benchmark dose estimation. *PeerJ*, 8, e10557.

```{r citation, eval=FALSE}
citation("bmd")
```


License
-------------------------

GPL-3. See the `LICENSE` file for details.

